Linear instability underlying synchronous waving of marine grass in flow

### 2014-09-21 Paper/grass.tex
Display equation in asymptotic analysis

\begin{abstract}
We show that the threshold flow condition for the spontaneous waving of marine grass arises due to the drag exerted by the vegetation on the flow. 
We show this by developing a self-consistent model of the flow past rigid vegetation. 
Linear stability analysis of the steady flow resulting from this model naturally leads to the threshold condition, and explains its mechanistic origin. 
Asymptotic analysis of the model for dense grass reveals two unstable modes. 
One of these modes shares many common features with the Kelvin-Helmholtz instability, but is demonstrated to be different from it. 
The predictions of the model for threshold condition and the resulting dominant frequency of waving agree with laboratory observations. 
\end{abstract}
\maketitle
%\section{Introduction}
Sea grasses exhibit a rich set of dynamical behavior due to their collective interaction with both steady and oscillatory flows.  
The resulting changes in hydrodynamic conditions can influence number of environmental processes such as transport of sediments, contaminants, dissolved oxygen, plant growth, and biomass production \cite{Fonseca87,Grizzle96,Nepf99,Nepf2012}. The simplest response of these flexible grass canopies to steady currents is the formation of coherent large amplitude oscillations, known as \monami \cite{AckermanOkubo93}. 
%Such motion is considered to play a central role in the recruitment 
%of microscopic marine organisms, such as blue mussel larvae\cite{Grizzle96}.
%The hydrodynamic mechanism underlying \monami is the focus of this paper. 
In this letter we provide a hydrodynamic mechanism for the onset of these coherent oscillations.
Current explanations of \monami invoke the existence of a shear layer near the top of the canopy \cite{Ghisal02,Raupach96} and its instability 
%due to different amounts of drag experienced by fluid in and above the canopy. 
%This shear layer is interpreted to become unstable to coherent vortices 
through a mechanism similar to the Kelvin-Helmholtz instability. This is thought to lead to coherent eddies over the sea grass canopy which then drives 
%Influence of these coherent eddies over sea grasses is manifested in their 
large amplitude synchronous oscillations.

@@

While the shear layer model appears to predict the frequency of \monami consistent with experimental observations, several aspect of existing theory remain unsatisfactory. 
(i) While the free shear layer is established due to the vegetation drag, linear stability analysis\cite{Raupach96} does not account for this, rendering this explanation inconsistent. 
(ii) Classical free shear flow is known to be unstable for all Reynolds number\cite{drazin}, whereas observations in the lab\cite{Ghisal02} and the field\cite{Grizzle96} indicate the existence of a threshold 
flow speed below which \monami is not observed. 
(iii)The thickness of the free shear layer, defined as the momentum thickness of the boundary layer near the grass tip, is in many cases comparable to the unvegetated layer thickness,
and therefore inconsistent with the assumptions of the free shear layer instability.
Here we present a self-consistent mathematical model that account correctly for these effects 
while also explaining lab experiments and field observations.
%Using this model, the synchronous waving may be understood to be arising from the linear hydrodynamic instability of a uniform flow over the grass bed, such that the principal role of the vegetation is to exert a drag on the flow. 
%A threshold flow speed arises naturally for the onset of the instability; this threshold condition and the resulting waving frequency compares well with experimental observations.
%The vegetation drag introduces a damping mechanism in the flow, in addition to the viscous damping, and overcoming this damping is the primary consideration that sets the threshold flow.
%We deduce this numerically and through an asymptotic scaling analysis in the limit of dense grass.
\newline
The drag exerted by the grass bed on the flow is central to the hypothesized mechanism\cite{Ghisal02} of the instability underlying \monami because it realizes the shear layer structure of the flow. 
The flexibility of the grass blades, on the other hand, serves merely aid in visualizing the flow structure resulting from the instability\cite{Nepf2012}. 
It has been demonstrated that the instability and the resulting flow structures are present even when the grass blades are replaced with rigid dowels\cite{Ghisal02}. 
Therefore, for simplicity, 
%mathematical model for the linear instability, 
we consider the grass bed to form a rigid structure that exchanges momentum with the surrounding fluid but does not deform. 
Indeed, a more detailed calculation undertaken for terrestrial grass demonstrated\cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, 
the flow instability is similar to the instability of a shear layer.

@@

To understand the onset of \monami, we first calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies 
\begin{equation}
 -{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0
\label{base_equ}
\end{equation}
where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$. In the limit of large grass density, shear stress exerted by the bottom surface is estimated to be negligible in comparison with the drag and hence the bottom surface ($y=0$) is approximated by zero shear.   
%The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, 
%as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{basicflow}. 
A comparison of laboratory mean profile with the steady flow profile is shown in figure \eqref{basicflow}.
Flow within grass has approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from the balance between drag and with pressure gradient. Flow outside has a simple parabolic velocity profile due to the balance between viscous forces and the pressure gradient. 
Continuity of flow speed and shear stresses at canopy top gives rise to a boundary layer 
%in the flow profile 
near the tip of the grass.

@@

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 2 has a threshold $\Rey \sim ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in figure \ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of \eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but $\Rey/\Ndg \sim O(1)$,.
Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the grass is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
This estimate simplifies \eqref{Orr-somerfield} within the grass to $\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi$, while in the unvegetated region it simplifies to the Rayleigh equation $ \left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$. 
This analysis predicts that the mode shape would converge in this limit, in agreement with our numerical results for Mode 2 shown in figure \ref{Asymptotic_mode}.

We interpret this as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the grass tip plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).

@@

This limit can be understood through asymptotically estimating the sizes of the terms in \eqref{Orr-somerfield} using $D\sim \delta^{-1}$, $\sigma \sim \delta^{-1}$, and $U\sim \delta$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. The terms balance when $\Rey \sim {\Ndg}^{1/2}$ (or $\Rey \sim H/\delta$), explaining the numerically observed asymptote (see figure \ref{Re_vs_delta}). This analysis predicts that the mode structure is self-similar over the length scale $\delta$; the verification of this expectation is shown in figure \ref{Asymptotic_mode}, supporting this argument.

>>>>

\begin{abstract}
We show that the threshold flow condition for the spontaneous waving of marine grass arises due to the drag exerted by the vegetation on the flow. 
We show this by developing a self-consistent model of the flow past rigid vegetation. 
Linear stability analysis of the steady flow resulting from this model naturally leads to the threshold condition, and explains its mechanistic origin. 
Asymptotic analysis of the model for dense grass reveals two unstable modes. 
One of these modes shares many common features with the Kelvin-Helmholtz instability, but is demonstrated to be different from it. 
The predictions of the model for threshold condition and the resulting dominant frequency of waving agree with laboratory observations. 
\end{abstract}
\maketitle
%\section{Introduction}
Sea grasses exhibit a rich set of dynamical behavior due to their collective interaction with both steady and oscillatory flows.  
The resulting changes in hydrodynamic conditions can influence number of environmental processes such as transport of sediments, contaminants, dissolved oxygen, plant growth, and biomass production \cite{Fonseca87,Grizzle96,Nepf99,Nepf2012}. 
The simplest response of these flexible grass canopies to steady currents is the formation of coherent large amplitude oscillations, known as \monami \cite{AckermanOkubo93}.  
In this letter we provide a hydrodynamic mechanism for the onset of these coherent oscillations.
Current explanations of \monami invoke the existence of a shear layer near the top of the canopy \cite{Ghisal02,Raupach96} and its instability through a mechanism similar to the Kelvin-Helmholtz instability. 
This is thought to lead to coherent eddies over the sea grass canopy which then drives large amplitude synchronous oscillations.

@@

While the shear layer model appears to predict the frequency of \monami consistent with experimental observations, several aspect of existing theory remain unsatisfactory. 
(i) While the free shear layer is established due to the vegetation drag, linear stability analysis\cite{Raupach96} does not account for this, rendering this explanation inconsistent. 
(ii) Classical free shear flow is known to be unstable for all Reynolds number\cite{drazin}, whereas observations in the lab\cite{Ghisal02} and the field\cite{Grizzle96} indicate the existence of a threshold 
flow speed below which \monami is not observed. 
(iii)The thickness of the free shear layer, defined as the momentum thickness of the boundary layer near the grass tip, is in many cases comparable to the unvegetated layer thickness,
and therefore inconsistent with the assumptions of the free shear layer instability.
Here we present a self-consistent mathematical model that account correctly for these effects 
while also explaining lab experiments and field observations.
%Using this model, the synchronous waving may be understood to be arising from the linear hydrodynamic instability of a uniform flow over the grass bed, such that the principal role of the vegetation is to exert a drag on the flow. 
%A threshold flow speed arises naturally for the onset of the instability; this threshold condition and the resulting waving frequency compares well with experimental observations.
%The vegetation drag introduces a damping mechanism in the flow, in addition to the viscous damping, and overcoming this damping is the primary consideration that sets the threshold flow.
%We deduce this numerically and through an asymptotic scaling analysis in the limit of dense grass.
\newline
The drag exerted by the grass bed on the flow is central to the hypothesized mechanism\cite{Ghisal02} of the instability underlying \monami because it realizes the shear layer structure of the flow. 
The flexibility of the grass blades, on the other hand, serves merely aid in visualizing the flow structure resulting from the instability\cite{Nepf2012}. 
It has been demonstrated that the instability and the resulting flow structures are present even when the grass blades are replaced with rigid dowels\cite{Ghisal02}. 
Therefore, for simplicity, we consider the grass bed to form a rigid structure that exchanges momentum with the surrounding fluid but does not deform. 
Indeed, a more detailed calculation undertaken for terrestrial grass demonstrated\cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, 
the flow instability is similar to the instability of a shear layer.

@@

To understand the onset of \monami, we first calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies 
\begin{equation}
 -{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0
\label{base_equ}
\end{equation}
where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$. 
In the limit of large grass density, shear stress exerted by the bottom surface is estimated to be negligible in comparison with the vegetation drag and hence the bottom surface ($y=0$) is approximated by zero shear.   
A comparison of laboratory mean profile with the steady flow profile is shown in figure \eqref{basicflow}.
Flow within grass has approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from the balance between drag and with pressure gradient. 
Flow outside has a simple parabolic velocity profile due to the balance between viscous forces and the pressure gradient. 
Continuity of flow speed and shear stresses at canopy top gives rise to a boundary layer near the tip of the grass.

@@

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 2 has a threshold $\Rey \sim ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in figure \ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of \eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but $\Rey/\Ndg \sim O(1)$,.
Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the grass is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
This estimate simplifies \eqref{Orr-somerfield} to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,   &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, &\text{ for $y>\hg$}.
\end{split}
\end{align}
Since $\Rey/\Ndg$ is the only remaining parameter, the mode shape would converge in the aforementioned limit, in agreement with our numerical results for Mode 2 shown in figure \ref{Asymptotic_mode}.

We interpret Mode 2 as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the grass tip plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).

@@

This limit can be understood through asymptotically estimating the sizes of the terms in \eqref{Orr-somerfield} using $D\sim \delta^{-1}$, $\sigma \sim \delta^{-1}$, and $U\sim \delta$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
The terms balance when $\Rey \sim {\Ndg}^{1/2}$ (or $\Rey \sim H/\delta$), leading to a simplification of \eqref{Orr-somerfield} as
\begin{align}
 \begin{split}
  \text{blah for $y<\hg$} \\
  \text{blah for $y>\hg$.}
 \end{split}
\end{align}
Since, $\Rey/\Ndg^{1/2}$ is the only remaining parameter in this model, explaining the numerically observed asymptote (see figure \ref{Re_vs_delta}). (this language should be improved)
This analysis concludes that the mode structure is self-similar over the length scale $\delta$; the verification of this expectation is shown in figure \ref{Asymptotic_mode}, supporting this argument.
